\documentclass[a4paper,10pt]{article}
\usepackage[american]{babel} % needed for iso dates
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{hyperref}
\usepackage{array}
\usepackage{underscore}
\usepackage{changepage}   % for the adjustwidth environment

\hypersetup{
  hidelinks
}

\newcommand{\todo}[1]{\emph{\textcolor{red}{TODO: #1}}}
\usepackage[top=2cm, bottom=2cm, left=2cm, right=2cm]{geometry}


\title{Proposal for the C++Now conference – Expected}
\author{Pierre T\textsc{albot} \and Vicente J. B\textsc{otet} E\textsc{scriba}}

\begin{document}

\maketitle
\todo{What should be the order of the authors? I alphabetically ordered it by the first name.\\
Several other name propositions:\\
The expected error monad.\\
Expected – An exceptional error monad.\\}
\section{Information}

\begin{itemize}
\item \textbf{Title} Expected — An exception-friendly error monad
\item \textbf{Type} Presentation.
\item \textbf{Optimum/minimum/maximum} 90/90/90 minutes.
\item \textbf{Level} Basic
\end{itemize}

\section{Abstract}



\section{Authors}

\subsection{Pierre T\textsc{albot}}

Pierre Talbot is a student at the University of Pierre et Marie Curie in Paris.

\subsection{Vicente J. B\textsc{otet} E\textsc{scriba}}

Vicente J. Botet Escriba got a Master in Computer Science from University Complutense of Madrid- Spain in 1986. More than 25 years of experience on software engineering for fault tolerant systems and more than 15 years of C++ experience. His main research area is on Concurrent Systems and Parallel Computing. Since January 2008 contributes actively to the Boost community. Co-author and maintainer of Boost.Thread/Chrono/Ratio. He is software engineer at Alcatel-Lucent France where he focuses on adaptation of very large telecom systems to multi-core platforms.

\section{Presentation overview}


\end{document}